\documentclass[a4paper,10pt,landscape,twocolumn]{scrartcl}

%% Settings
\newcommand\problemset{1}
\newcommand\deadline{Wednesday, 7 November 2018, 12:00h}
\newif\ifcomments
\commentsfalse % hide comments
%\commentstrue % show comments

% Packages
\usepackage[english]{exercises}
\usepackage{wasysym}
\usepackage{hyperref}
\hypersetup{colorlinks=true, urlcolor = blue, linkcolor = blue}

\usepackage{tikz}

\begin{document}

\homeworkproblems

\begin{exercise}[Deriving the weak law of large numbers (9pt)]
	\begin{subex}[(2pt)] (Markov's inequality) For any real non-negative random variable $X$ and any $t > 0$, show that
	\[
	P_X(X \geq t) \leq \frac{\mathbb{E}[X]}{t}\, .
	\]
\end{subex}
\begin{subex}[(1pt)]
	Exhibit a random variable (which can depend on $t$) that achieves this inequality with equality.
	\end{subex}

	\begin{subex}[(3pt)] (Chebyshev's inequality.) Let $Y$ be a random variable with mean $\mu$ and variance $\sigma^2$. Show that for any $\varepsilon > 0$,
	\[
	P(|Y - \mu| \geq \varepsilon) \leq \frac{\sigma^2}{\varepsilon^2} \, .
	\]
	\textbf{Hint:} Define a random variable $X := (Y - \mu)^2$.
	\end{subex}
	\begin{subex}[(3pt)] (The weak law of large numbers.) Let $Z_1, Z_2, ..., Z_n$ real i.i.d. random variables with mean $\mu = \mathbb{E}[Z_i]$ and variance $\sigma^2 = \mathbb{E}[(X_i - \mu)^2] < \infty$. Define the random variables $S_n = \frac{1}{n} \sum_{i=1}^n Z_i$. Show that
	\[
	P(|S_n - \mu | \geq \varepsilon) \leq \frac{\sigma^2}{n\varepsilon^2}\, .
	\]
	Thus, $P(|S_n - \mu| \geq \varepsilon) \to 0$ as $n \to \infty$. This is known as the weak law of large numbers (which we will use heavily in Week 03).
	\end{subex}
\end{exercise}

\begin{exercise}[Multiple-choice test (3pt)]
A multiple-choice exam has 4 choices for each question. A student has studied enough so that the probability she will know the answer to a question is 0.5, the probability that she will be able to eliminate one choice is 0.25, otherwise all 4 choices seem equally plausible. If she knows the answer she will get the question right. If not she has to guess from the 3 or 4 choices.
As the teacher you want the test to measure what the student knows. If the student answers a question correctly what’s the probability she knew the answer?
\end{exercise}

\begin{exercise}[Probability urn (3pt)]
There are 7 white and 4 black balls in a urn. You draw three balls in sequence at random. On each draw, if the ball is white you set it aside and if the ball is black you put it back in the urn. What is the probability that the third draw is white? (If you get a black ball it counts as a draw even though you put it back in the urn)
\end{exercise}

\begin{exercise}[Birthday party (9pt)]
 20 ILLC members are having a party.

\begin{subex}[(3pts)]
To prepare, they need to choose 3 people to set the table, 2 people to bake cake and 6 people to clean up. Each person can only do 1 task (this doesn't add up to 20, the rest of the people don't help). In how many different ways can they choose which people perform these tasks?
\end{subex}
\begin{subex}[(3pts)]
The party crowd consists of 5 staff members and 15 students, and tea and coffee is served after the cake. It turns out that all staff members don't like tea. If they only give tea to 10 of the 20 people, what is the probability that only students get tea?
\end{subex}
\begin{subex}[(3pts)]
If they only give tea to 10 of the 20 people, what is the probability that 9 students and 1 staff member gets tea?
\end{subex}
\end{exercise}

\begin{exercise}[Expectation and variance (4pt)]
	\begin{subex}[(2pt)]
	Show that expectation is linear: for arbitrary $a,b \in \mathbb{R}$ and arbitrary real random variables $X,Y$,
	\[
	\mathbb{E}[aX + bY] = a\mathbb{E}[X] + b \mathbb{E}[Y].
	\]
	\end{subex}
	\begin{subex}[(2pt)]
	Use Jensen's inequality to show that $\mbox{Var}(X) \geq 0$ for any $X$.
	\end{subex}
\end{exercise}


\begin{exercise}[Two coins and a die (8pt)]
  You have two (fair) coins and a (fair) 4-sided die with outcomes $\{1,2,3,4\}$. Let $X$ be the number of
  heads after flipping the two coins and let $Y$ be the result of rolling the
  die. Let $Z$ be the average of $X$ and $Y$.
\begin{subex}[(2pt)]
What is the distribution $P_Z$ of $Z$?
\end{subex}
  \begin{subex}[(3pt)]
    Compute the variations of $X$, $Y$ and $Z$.	
  \end{subex}
  \begin{subex}[(3pt)]
    You play the following game. If $2X \ge Y$, you win $X^2$ euros and
    otherwise you lose 1 euro. What is your expected total gain or loss after
    playing this game $40$ times?
  \end{subex}
\end{exercise}


\begin{exercise}[Fermi estimates (8pt)]
Read up on what a
\href{https://en.wikipedia.org/wiki/Fermi_problem}{Fermi estimate}
is. Give a Fermi estimate for the \emph{daily number of git commits
in the world}.
 
In this exercise, it is more important to be clear about your
arguments and assumptions rather than getting the correct result.

\begin{subex}[(4pt)]
Give your Fermi estimate without looking up any numbers on the internet.
\end{subex}

\begin{subex}[(4pt)]
Look up the number for your estimates from the previous
sub-exercise. Possibly revise your strategy to do the
estimate. Compare the new final outcome to what you got previously as
endresult in (a).
\end{subex}

One team member should post your answers to
\href{https://canvas.uva.nl/courses/2205/discussion_topics/23268}{this
  discussion forum}, don't forget to state your team name.
\end{exercise}

\begin{exercise}[Programming... (5pt)]
For a real parameter $0 < p < \infty$, the $L_p$-distance between two probability distributions $P$ and $Q$
over the finite alphabet $\mathcal{X}$ is defined as 
\begin{align*}
\| P - Q \|_p := \left( \sum_{x \in \mathcal{X} } | P(x) - Q(x) |^p \right) ^{(1/p)}
\end{align*}
For $p=1$, the value $\frac12 \| P - Q \|_1$ is called the
\emph{statistical distance} or \emph{total variation distance} 
  between $P$ and $Q$.

In this exercise, we are going to analyze the letter frequencies of \emph{Alice in
  Wonderland} in five different languages: English, German, Esperanto,
Italian, Finnish. Here, we are going to ignore that languages like German and Finnish have important other letters such as {\"a}, {\"o}, {\"u}, and only consider the 26 English letters and space.

\newcommand{\lang}{\textrm{lang}}
For $\lang \in \{ \textrm{eng, ger, esp, ita, fin} \}$, let $P_{\lang}$ be the frequency distribution of the 26 English letters and space of Alice in Wonderland.

\begin{subex}
Compute $\| P_{\textrm{eng}} - P_{\textrm{ger}} \|_p$ for $p \in \{0.01, 0.2, 0.5, 1, 1.5, 10, 1000\}$. Plot the $L_p$ distance as function of $0 < p < \infty$.
\end{subex}

\begin{subex}
For $p=1$, compute all pairwise distances $\| P_{\lang} - P_{\lang'} \|_1$ for $\lang, \lang' \in \{ \textrm{eng, ger, esp, ita, fin} \}$.
\end{subex}

\begin{subex}

\end{subex}

\end{exercise}




\end{document}