\documentclass[a4paper,10pt,landscape,twocolumn]{scrartcl}

%% Settings
\newcommand\problemset{4}
\newcommand\deadline{Wednesday, 28 November 2018, 12:00h}
\newif\ifcomments
\commentsfalse % hide comments
%\commentstrue % show comments

% Packages
\usepackage[english]{exercises}
\usepackage{wasysym}
\usepackage{hyperref}
\hypersetup{colorlinks=true, urlcolor = blue, linkcolor = blue}

\usepackage{tikz}

\begin{document}

\homeworkproblems

{\sffamily\noindent
Unless otherwise stated, you should provide exact answers rather than rounded numbers (e.g., $\log 3$ instead of 1.585) for non-programming exercises.
}

\newcommand{\typsetA}{A^{(n)}_{\varepsilon}}
\newcommand{\typsetB}{B^{(n)}_{\delta}}

\begin{exercise}[Calculation of the typical set (7pt)]
To clarify the notion of a typical set $\typsetA$ and the smallest set of high probability $\typsetB$, we will calculate these sets for a simple example. Consider a sequence of i.i.d.\ binary random variables $X_1, X_2, . . . X_n$, where the probability that $P_X(1) = 0.6$ and $P_X(0) = 0.4$.
	\begin{subex}[(3pt)]
	With $n = 25$ and $\varepsilon = 0.1$, which sequences fall in the typical set $\typsetA$? What is the probability of the typical set (three decimals)? How many elements are there in the typical set? (This involves computation of a table of probabilities for sequences with $k$ 1's, $0 \leq k \leq 25$, and finding those sequences that are in the typical set.)
\\\textbf{Hint:} Here is the table: \url{http://goo.gl/sQCPM0}
	\end{subex}
	\begin{subex}[(2pt)]
	How many elements are there in the smallest set that has probability 0.9? In other words, what is $|\typsetB|$ for $n = 25$ and $\delta = 0.1$?
	\end{subex}
	\begin{subex}[(2pt)]
	How many elements are there in the intersection $|\typsetA \cap \typsetB|$ of the sets computed in parts (b) and (c)? What is the probability of this intersection (three decimals)?
	\end{subex}
\end{exercise}

\begin{exercise}[Three random variables (3pt)]
Let $A,B,C$ be random variables such that
\begin{align*}
I(A;B) &= 0\\
I(A;C|B) &= I(A;B|C)\\
H(A|BC) &= 0
\end{align*}
Prove one of the following possible relations between $H(A)$ and $H(C)$: $=$, $\leq$, $\geq$, $<$, or $>$. If you show one of the weaker inequalities ($\leq$ or $\geq$), also exhibit an example that shows why the inequality is not strict.
\end{exercise}

\begin{exercise}[Piece of cake (3pt)]
A big cake is repeatedly sliced into two pieces. At every step, the \emph{smaller} part is discarded (or eaten). On the other part the process is continued. At every step, the remaining piece is cut randomly into two pieces with the following proportions:
\begin{align*}
\left(\frac{2}{3},\frac{1}{3}\right) &\mbox{ with probability } \frac{3}{4},\\
\left(\frac{3}{5},\frac{2}{5}\right) &\mbox{ with probability } \frac{1}{4}.\\
\end{align*}
For example, three consecutive cuts might result in a piece of cake of size $\frac{3}{5} \cdot \frac{2}{3} \cdot \frac{2}{3}$. Let $T_n$ be the fraction of the cake left after $n$ cuts. Clearly, this fraction $T_n$ decreases exponentially with $n$, i.e. $T_n \approx c^n$ for some real constant $0<c<1$. Determine $c$ for large values of $n$.

\textbf{Hint:} Let $C_i$ be a random variable describing the fraction of the cake that is cut (and kept) at the $i$th cut, i.e.\ $C_i = \frac{2}{3}$ with probability $\frac{3}{4}$, or $\frac{3}{5}$ otherwise.  Then use the weak law of large numbers.





\end{exercise}


\begin{exercise}[Programming (7pt)]
	In one of the quizzes for this week, you implemented the encryption and decryption procedure for the Vigen\`{e}re cipher. (Note: we treat ``\texttt{A}" as 1 instead of 0. That is, \texttt{K+A=L})
	\begin{subex}[(1pt)]
		Consider this cipher for a fixed message length $m$ and a fixed key length $k$.
		Suppose the possible messages are uniformly distributed over all monocase messages (excluding spaces; there are 26 possible letters) of length $m$. What is $I(M;C)$? When is the Vigen\`ere cipher perfectly secure (i.e., $I(M;C) = 0$)?
	\end{subex}
	In reality, the possible messages are not uniformly distributed. We can use this fact to break a ciphertext that was encoded with the Vigen\`{e}re cipher.
	
	Recall the definition of collision probability from the first homework:
	\[
	Coll(P) := \sum_{x \in \mathcal{X}} P(x)^2.
	\]
	Let $Coll(P_M)$, $Coll(P_K)$, and $Coll(P_C)$ denote the collision probabilities of (sampling a single letter form) the plaintext, key, and ciphertext, respectively. Suppose you know that the message is in English. In the first homework, you estimated that $Coll(P_M) \approx 0.0655$.
	\begin{subex}[(2pt)]
		Show that
		\[
		Coll(P_C) \approx \frac{1}{k} Coll(P_M) + \frac{k-1}{k} Coll(P_K)
		\]
		by assuming that:
		\begin{enumerate}
			\item the message is very long compared to the key ($m$ is very large), and
			\item the letters in the message are i.i.d. according to $P_M$.
			\item the letters in the key are i.i.d. according to a uniform $P_K$.
		\end{enumerate}
	\end{subex}
	\begin{subex}[(2pt)]
		Use (b) to find the most likely key length $k$ in the following ciphertext:
		\begin{center}
			\texttt{QMIXFFKFMVPZSVYGOMAPOTTCUUBGIIPPOTPZEVNVJHIUBGSCDCIPXBKDBR
			SIXFMVIUVJBCWYNPMVNSDFNDZDWQCVFLKZCGMXYJNJBAQPXISUEFVGGIT
			FMQNOEIATBWG}
		\end{center}
	\end{subex}
	\begin{subex}[(2pt)]
		The ciphertext in the previous subexercise was sent to Cleopatra by Caesar. That means it will very likely contain one or more of the following words: \texttt{CLEOPATRA}, \texttt{CAESAR}, \texttt{QUEEN}, \texttt{GRACE}, \texttt{ALEXANDRIA}. Use this side information, in combination with the key length you computed in (c), to figure out the original message. What was the encryption key?
	\end{subex}
\end{exercise}



\end{document}